The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics

### 2016-01-31 Papers/InProgress/IJCAI/submission.tex
Improve section 2

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz. 
%\cite{ToDo:Leibniz}.  
Although G\"odel's notion of positive property is not exactly the same as Leibniz's notion of perfection, G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. The main difference to G\"odel's version is an extra conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen [\citeyear[p.365]{Hazen}] states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
However, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5U}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  at the time that G\"odel's axioms were inconsistent.}

@@

For more than four decades, the serious consequences of G\"odel's
mistake remained unnoticed, despite intense activity on
criticising his argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they must be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

>>>>

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz. 
%\cite{ToDo:Leibniz}.  
Although G\"odel's notion of positive property is not exactly the same as Leibniz's notion of perfection, G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. The main difference to G\"odel's version is an extra conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

G\"odel's omission has been
considered inessential and merely an oversight by many. For instance,
Hazen [\citeyear[p.365]{Hazen}] states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they must be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

However, as explained here, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5U}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  at the time that G\"odel's axioms were inconsistent.}
